---
title: "Project_0_Biologist"
author: "Camilla Belamarich"
date: "2/22/2021"
output: html_document
---
Load DE matrix and change the first column "X" to "probe_id"
```{r}
de_matrix <- read.csv("~/Desktop/BF528/BF528/4_5_Welch", header = TRUE)
names(de_matrix)[names(de_matrix) == "X"] <- "probe_id"
```
Create probe set ids as character vector to plug into select function and map gene symbols to DE matrix
```{r}
library(hgu133plus2.db)
pset_ids <- de_matrix$probe_id
pset_ids <- as.character(pset_ids)
mapped_results <- select(hgu133plus2.db, keys=pset_ids, columns=("SYMBOL"))
```
Adding mapped gene symbols to DE matrix
```{r}
new_de_matrix <- merge(x=de_matrix, y=mapped_results, by.x="probe_id", by.y="PROBEID")
summary(new_de_matrix)
```
NA values were found in matrix, so they have to be removed before moving forward.
```{r}
new_de_matrix <- new_de_matrix[!is.na(new_de_matrix$SYMBOL),]
summary(new_de_matrix)
```
Multiple Probeset IDs matched to the same gene symbol. In the paper, researchers picked the probeset with the greatest variance out of the ones mapped to a single gene symbol.
```{r}
gene_symbols <- new_de_matrix$SYMBOL
length(gene_symbols)
unique_gene_symbols <- unique(gene_symbols)
length(unique_gene_symbols)
```
Removing duplicates with the greatest variance
```{r}
library(tibble)
library(dplyr)
# for symbol
new_de_matrix <- as_tibble(new_de_matrix)
removed_duplicates <- new_de_matrix %>%
  group_by(SYMBOL) %>%
  filter(p_adj == max(p_adj)) %>%
  ungroup(SYMBOL)
summary(removed_duplicates)
```

Selecting the top 1000 up- and down-regulated genes
```{r}
# Up-regulated
up <- slice_max(removed_duplicates, order_by=removed_duplicates$t_test_statistic, n=1000)
up_regulated <- slice_max(removed_duplicates,order_by=removed_duplicates$t_test_statistic, n=10)
# Down-regulated
down <- slice_min(removed_duplicates, order_by=removed_duplicates$t_test_statistic, n=1000)
down_regulated <- slice_min(removed_duplicates,order_by=removed_duplicates$t_test_statistic, n=10)
```
Export DE genes
```{r}
write.csv(up_regulated, "~/Desktop/BF528/BF528/up_regulated.csv", row.names = FALSE)
write.csv(down_regulated, "~/Desktop/BF528/BF528/down_regulated.csv", row.names = FALSE)
```

Read in Kegg, GO, and Hallmark gene sets from MiSigDB with GSEABase
```{r}
library(GSEABase)
kegg <- getGmt("c2.cp.kegg.v7.2.symbols.gmt")
n_kegg <- length(kegg) # number of kegg gene sets
go <- getGmt("c5.all.v7.2.symbols.gmt")
n_go <- length(go) # number of go gene sets
hallmark <- getGmt("h.all.v7.2.symbols.gmt")
n_hallmark <- length(hallmark) # number of hallmark gene sets
```
Reporting number of gene sets in each:
```{r}
print("Number of gene sets in KEGG:")
print(n_kegg)
print("Number of gene sets in G0:")
print(n_go)
print("Number of gene sets in Hallmark:")
print(n_hallmark)
```
[1] "Number of gene sets in KEGG:"
[1] 186
[1] "Number of gene sets in G0:"
[1] 14765
[1] "Number of gene sets in Hallmark:"
[1] 50


Selecting the statistically significant deferentially expressed genes using the adjusted p-value:
```{r}
DE <- removed_duplicates[removed_duplicates$p_adj < 0.05,]
DE_symbols <- DE$SYMBOL
all_gene_symbols <- removed_duplicates$SYMBOL
non_DE_symbols <- setdiff(all_gene_symbols, DE_symbols)
```
Printing number of each group to prepare for contigency tables:
```{r}
print("Total number of DE genes:")
length(DE_symbols)
print("Total number of non DE genes:")
length(non_DE_symbols)
print("Total number of genes left after removing duplicates:")
length(all_gene_symbols)
```
[1] "Total number of DE genes:"
[1] 8663
[1] "Total number of non DE genes:"
[1] 8623
[1] "Total number of genes left after removing duplicates:"
[1] 17287


Contingency table needs:
-total number of differentially expressed genes
-total number of not differentially expressed genes
# these are the only inputs needed for function
-number of differentially expressed genes in the gene set
-number of not differentially expressed genes in the gene set
-number of differentially expressed genes not in the gene set
-number of not differentially expressed genes not in the gene set
#
-total number of genes in the gene set
-total number of genes not in the gene set
-overall total genes

Function that accepts a gene list and a single gene set:
```{r}
contingency_function <- function(gene_set, de_gene_set, non_de_gene_set){
  # total number of differentially expressed genes
  n_differentially_expressed <- length(geneIds(de_gene_set))
  # total number of not differentially expressed genes
  n_not_differentially_expressed <- length(geneIds(non_de_gene_set))
  # number of differentially expressed genes in the gene set
  differentially_expressed_in <- gene_set & de_gene_set
  n_differentially_expressed_in <- length(geneIds(differentially_expressed_in))
  # number of differentially expressed genes not in the gene set
  n_differentially_expressed_not_in <- n_differentially_expressed - n_differentially_expressed_in
  # number of not differentially expressed genes in the gene set
  not_differentially_expressed_in <- gene_set & non_de_gene_set
  n_not_differentially_expressed_in <- length(geneIds(not_differentially_expressed_in))
  # number of not differentially expressed genes not in the gene set
  n_not_differentially_expressed_not_in <- n_not_differentially_expressed - n_not_differentially_expressed_in
  contingency_table <- c(n_differentially_expressed_in, n_differentially_expressed_not_in, n_not_differentially_expressed_in, n_not_differentially_expressed_not_in)
  return(contingency_table)
}
```


GSEABase gene set inputs:
```{r}
gene_set_de <- GeneSet(DE_symbols, setName = "differentially expressed")
gene_set_not_de <- GeneSet(non_DE_symbols, setName = "not differentially expressed")
```

Contingency tables:

KEGG
```{r}
contingency_kegg <- list()
for (i in kegg) {
  contigency <- contingency_function(i, gene_set_de, gene_set_not_de)
  contingency_kegg[[setName(i)]] <- contigency
}
```
GO
```{r}
contingency_go <- list()
for (i in go) {
    contigency <- contingency_function(i, gene_set_de, gene_set_not_de)
  contingency_go[[setName(i)]] <- contigency
}
```
HALLMARK
```{r}
contingency_hallmark <- list()
for (i in hallmark) {
    contigency <- contingency_function(i, gene_set_de, gene_set_not_de)
  contingency_hallmark[[setName(i)]] <- contigency
}
```


Fisher data frames:
```{r}
kegg_fisher_test <- data.frame(gene_set = character(), statistic_estimate = double(), p_value = double())

go_fisher_test <- data.frame(gene_set = character(), statistic_estimate = double(), p_value = double())

hallmark_fisher_test <- data.frame(gene_set = character(), statistic_estimate = double(), p_value = double())
```
Fisher test results:
```{r}
for (i in names(contingency_kegg)) {
  fisher_test <- fisher.test(matrix(contingency_kegg[[i]], nrow = 2))
  gene_set = i
  statistic_estimate <- fisher_test$estimate[[1]]
  p_value <- fisher_test$p.value
  kegg_fisher_test[nrow(kegg_fisher_test) + 1,] <- c(i, statistic_estimate, p_value)
}
kegg_gene_set_name <- names(contingency_kegg)
kegg_fisher_test$gene_set <- kegg_gene_set_name
kegg_fisher_test
```

```{r}
for (i in names(contingency_go)) {
  fisher_test <- fisher.test(matrix(contingency_go[[i]], nrow=2))
  gene_set = i
  statistic_estimate <- fisher_test$estimate[[1]]
  p_value <- fisher_test$p.value
  go_fisher_test[nrow(go_fisher_test) + 1,] <- c(i, statistic_estimate, p_value)
}
go_gene_set_name <- names(contingency_go)
go_fisher_test$gene_set <- go_gene_set_name
go_fisher_test
```

```{r}
for (i in names(contingency_hallmark)) {
  fisher_test <- fisher.test(matrix(contingency_hallmark[[i]], nrow=2))
  gene_set = i
  statistic_estimate <- fisher_test$estimate[[1]]
  p_value <- fisher_test$p.value
  hallmark_fisher_test[nrow(hallmark_fisher_test) + 1,] <- c(i, statistic_estimate, p_value)
}
hallmark_gene_set_name <- names(contingency_hallmark)
hallmark_fisher_test$gene_set <- hallmark_gene_set_name
hallmark_fisher_test
```

Adjust the p-values for multiple hypotheses using the Benjamini-Hochberg (FDR) procedure and append this adjusted p-value column to the data frame.
```{r}
kegg_fisher_test$pvalue_adjusted <- p.adjust(kegg_fisher_test$p_value, method = 'BH')
go_fisher_test$pvalue_adjusted <- p.adjust(go_fisher_test$p_value, method = 'BH')
hallmark_fisher_test$pvalue_adjusted <- p.adjust(hallmark_fisher_test$p_value, method = 'BH')
```

the number of significantly enriched gene sets at adjusted ð‘<0.05:
```{r}
kegg_significant <- kegg_fisher_test[kegg_fisher_test$pvalue_adjusted<0.05,]
go_significant <- go_fisher_test[go_fisher_test$pvalue_adjusted<0.05,]
hallmark_significant <- hallmark_fisher_test[hallmark_fisher_test$pvalue_adjusted<0.05,]

n_kegg_significant <- length(kegg_significant$gene_set)
n_go_significant <- length(go_significant$gene_set)
n_hallmark_significant <- length(hallmark_significant$gene_set)

print("Number of significantly enriched gene sets in KEGG:")
print(n_kegg_significant)
print("Number of significantly enriched gene sets in GO:")
print(n_go_significant)
print("Number of significantly enriched gene sets in Hallmark:")
print(n_hallmark)
print("Total number of significantly enriched genes:")
print(n_kegg_significant + n_go_significant + n_hallmark_significant)
```
[1] "Number of significantly enriched gene sets in KEGG:"
[1] 17
[1] "Number of significantly enriched gene sets in GO:"
[1] 246
[1] "Number of significantly enriched gene sets in Hallmark:"
[1] 50
[1] "Total number of significantly enriched genes:"
[1] 275

Sort each dataframe by nominal p-value and report the top three results for each in a table in your report. 
```{r}
top3_kegg <- slice_min(kegg_fisher_test, order_by = pvalue_adjusted, n=3)
top3_go <- slice_min(go_fisher_test, order_by = pvalue_adjusted, n=3)
top3_hallmark <- slice_min(hallmark_fisher_test, order_by = pvalue_adjusted, n=3)

top3_results <- rbind(top3_kegg, top3_go, top3_hallmark)

top3_results

write.csv(top3_results, "~/Desktop/BF528/BF528/top3_results.csv", row.names = FALSE)
```


